|

### 2024-08-29 reproducibility/manuscript/manuscript.qmd
updated and tersened analytic solutions methods sub-section

We assume the dynamical gene expression is determined by the RNA splicing
process, and infer the unspliced and spliced gene expression level from the
ordinary differential equation (ODEs) proposed in velocyto [@La_Manno2018-lj] and scVelo
[@Bergen2020-pj]
\begin{align}
\frac{du}{dt} &= \alpha(t) - \beta u, \quad u(0) = u_0 \label{eq-dudt}\\
   \frac{ds}{dt} &= \beta u - \gamma s, \quad s(0) = s_0, \label{eq-dsdt}
\end{align}
where $u(t), s(t)$ are the unspliced and spliced expression levels of a gene at time $t$ under a transcription rate $\alpha(t)$ with possible temporal dependence, splicing rate $\beta$, and degradation rate $\gamma$. We specify this model to a setting that depends on cell $c$ and gene $g$ as follows:
\begin{align}
\frac{du_{cg}}{dt} &= \alpha_{cg}(t) - \beta_{g} u_{cg}, \quad u_{cg}(0) = u_{cg}^{(0)} \label{eq-dudt}\\
   \frac{ds_{cg}}{dt} &= \beta_{g} u_{cg} - \gamma_{g} s_{cg}, \quad s_{cg}(0) = s_{cg}^{(0)} \label{eq-dsdt}.
\end{align}
In the equation, the subscript $c$ is the cell dimension, $g$ is the gene dimension,
$\left( u_{cg}(t), s_{cg}(t) \right)$ 
are the unspliced and spliced expression functions given the
change of time per cell and gene.
We restrict attention to piecewise-constant $\alpha_{cg}(t)$ to capture gene-specific activation and repression. We take special care to model a gene- and cell-specific switching time that marks a single transition from activation to repression by introducing a Bernoulli variable $k_{cg}$ to model unknown activation state. We assume our cell-by-gene data-matrix arrive as observations of Poisson-counts related to the solution of the above ODEs at unknown times $\tau_{cg}$, which is modeled as a relationship between an unknown latent time shared across each cell, $t_c$, and unknown gene-specific time-offsets $t_{0,g}$ where all read counts for a single cell occurred at an unknown, but shared latent time $t_c$. These relative times are also used to parametrize the Bernoulli process for $k_{cg}$. 
Importantly, we recognize that the initial conditions are in fact unknown.

@@

Next we introduce the priors for the splicing parameters (where the activation rate $\alpha$ depends on the activation state $k_{cg}$ from above):
\begin{align}
  \alpha^{(0)}_g &\sim \text{LogNormal}(0, 1) \\
  \beta_g &\sim \text{LogNormal}(0, 1) \\
  \gamma_g &\sim \text{LogNormal}(0, 1) \\
  \alpha_{cg} &= \begin{cases}
    \alpha^{(0)}_g & \text{if } k_{cg} = 0 \\
    0 & \text{if } k_{cg} = 1
  \end{cases}
\end{align}
\textbf{Note that $\alpha^{(1)}$ is shared for all the genes, while ${\alpha^{(0)}}_g$ is learned independently for
each gene. MATT: this was in the old text, but I think $\alpha^{(1)}$ is no longer used based on conversations with Alvin?}

@@

Next, we define the observation model that gives rise to the observed counts as:
\begin{align}
  \mu^{(u)}_c &= \sum_{g=1}^G {u}^{\text{(obs)}}_{cg}, \quad \mu^{(s)}_c = \sum_{g=1}^G {s}^{\text{(obs)}}_{cg} \\
  \sigma^{(u)}_c &= \sqrt{\frac{1}{G} \sum_{g=1}^G \left( u_{cg}^{\text{(obs)}} - \mu^{(u)}_c \right)^2} \\
  \sigma^{(s)}_c &= \sqrt{\frac{1}{G} \sum_{g=1}^G \left( s_{cg}^{\text{(obs)}} - \mu^{(s)}_c \right)^2} \\
  \eta^{(u)}_c &\sim \text{Normal}\Big(\mu^{(u)}_c, \ \sigma^{(u)}_c\Big) \\
  \eta^{(s)}_c &\sim \text{Normal}\Big(\mu^{(s)}_c, \ \sigma^{(s)}_c\Big) \\
  \hat{\mu}^{(u)}_c &= \sum_{g=1}^G \hat{u}_{cg}, \quad \hat{\mu}^{(s)}_c = \sum_{g=1}^G \hat{s}_{cg} \\
  \lambda^{(u)}_{cg} &= \log(\hat{u}_{cg}) + \log(\eta^{(u)}_{c}) - \log(\hat{\mu}^{(u)}_c) \\
  \lambda^{(s)}_{cg} &= \log(\hat{s}_{cg}) + \log(\eta^{(s)}_{c}) - \log(\hat{\mu}^{(s)}_c) \\
  \hat{u}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(u)}_{cg})\Big) \\
  \hat{s}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(s)}_{cg})\Big)
\end{align}
Here, we use ${u}^{\text{(obs)}}_{cg}, {s}^{\text{(obs)}}_{cg}$ to denote the observed unspliced and spliced counts
for cell $c$ and gene $g$. We use $\hat{u}^{\text{(obs)}}_{cg}, \hat{s}^{\text{(obs)}}_{cg}$ to
denote our generative model's prediction of these unspliced and spliced expression levels.
The generative process for modeling these observed read counts given denoised gene transcript expression level $\hat{u}_{cg}, \hat{s}_{cg}$ considers the expected number of observed reads for a given gene in a given cell as the number of transcripts times the ratio of the cell's total reads to total transcripts.
\textbf{Improve descriptions of how noise is modeled in the observation model.}

\textbf{Need to update the analytic solutions, but first need to confirm the above is correct. Also, I recommend pushing all of the below analytic solutions to the appendix.}
The analytic solution of the differential equations to predict
spliced and unspliced gene expression given their parameters is derived by the
authors of scVelo and a theoretical RNA velocity study
[@Bergen2020-pj;@Li2021-qa] and given in
Eqs. \ref{eq-solution-u}-\ref{eq-solution-s2}.
\begin{align}
u\left(\tau^{\left(k_{c g}\right)}\right) 
  &= u_0^{\left(k_{c g}\right)}{ }_g e^{-\beta_g \tau^{\left(k_{c g}\right)}} 
  \nonumber \\
&\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}}
  {\beta_g}\left(1-e^{-\beta_g \tau^{\left(k_{c g}\right)}}\right) 
  \label{eq-solution-u}\\
s\left(\tau^{\left(k_{c g}\right)}\right) 
  &= s_0^{\left(k_{c g}\right)} e^{-\gamma_g \tau^{\left(k_{c g}\right)}} 
  \nonumber \\
  &\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}}{\gamma_g}
    \left(1-e^{-\gamma_g \tau^{\left(k_{c g}\right)}}\right) 
    \nonumber\\
  &\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}-\beta_g u_0^{\left(k_{c g}\right)}}
    {\gamma_g-\beta_g}\left(e^{-\gamma_g \tau^{\left(k_{c g}\right)}}
    -e^{-\beta_g \tau^{\left(k_{c g}\right)}}\right), 
    \nonumber \\
  &\qquad \beta \neq \gamma \label{eq-solution-s} \\
s\left(\tau^{\left(k_{c g}\right)}\right) 
  &= {s_0^{\left(k_{c g}\right)}}_g e^{-\beta_g \tau^{\left(k_{c g}\right)}} \nonumber \\
  &\hskip -24pt +\frac{\alpha^{\left(k_{c g}\right)}}{\beta_g}
  \left(1-e^{-\beta_g \tau^{(k c g)}}\right) 
  \nonumber \\
  &\hskip -24pt -\left(\alpha^{\left(k_{c g}\right)}
  -\beta_g u_0^{\left(k_{c g}\right)}{ }_g\right) \tau^{\left(k_{c g}\right)} 
  e^{-\beta_g \tau^{\left(k_{c g}\right)}}, \nonumber \\
  &\qquad \beta = \gamma \label{eq-solution-s2}
\end{align}

To simplify these equations, consider the case when $k_{cg} = 0$ and 
$\beta_g \neq \gamma_g$. Then, 
\begin{align}
u\left(\tau^{(0)}\right) &= {u_0^{(0)}}_g e^{-\beta_g \tau^{(0)}} \nonumber \\
  & \hskip -24pt + \frac{{\alpha^{(0)}}_g}{\beta_g}\left(1-e^{-\beta_g \tau^{(0)}}\right), 
  \label{eq-sol-usimp} \\
s\left(\tau^{(0)}\right) &= s_0^{(0)}{ }_g e^{-\gamma_g \tau^{(0)}} \nonumber \\
  & \hskip -24pt +\frac{{\alpha^{(0)}}_g}{\gamma_g}\left(1-e^{-\gamma_g \tau^{(0)}}\right) 
  \nonumber\\
  & \hskip -24pt +\frac{{\alpha^{(0)}}_g-\beta_g {u_0^{(0)}}_g}{\gamma_g-\beta_g}
  \left(e^{-\gamma_g \tau^{(0)}}-e^{-\beta_g \tau^{(0)}}\right). \label{eq-sol-ssimp}
\end{align}
When $k_{cg} = 0$ and $\beta_g = \gamma_g$, then $u\left(\tau^{(0)}\right)$ 
has the same solution, and $s\left(\tau^{(0)}\right)$ becomes
\begin{align}
s\left(\tau^{(0)}\right) &= s_0^{(0)}{ }_g e^{-\gamma_g \tau^{(0)}} \nonumber \\
  & \hskip -24pt +\frac{{\alpha^{(0)}}_g}{\gamma_g}
    \left(1-e^{-\gamma_g \tau^{(0)}}\right) \nonumber\\
  & \hskip -24pt - \left( {\alpha^{(0)}}_g-\beta_g {u_0^{(0)}}_g \right) 
    \tau^{(0)} e^{-\beta_g \tau^{(0)}}. \label{eq-sol-ssimp2}
\end{align}
When $k_{cg} = 1$ and $\beta_g \neq \gamma_g$, then
\begin{align}
u\left(\tau^{(1)}\right) &=u_0^{(1)} g^{e^{-\beta_g \tau^{(1)}}}, \\
s\left(\tau^{(1)}\right) &=s_0^{(1)} e^{-\gamma_g \tau^{(1)}} \nonumber \\
  & \hskip -24pt +\frac{-\beta_g u_0^{(1)}}{\gamma_g-\beta_g}
  \left(e^{-\gamma_g \tau^{(1)}}-e^{-\beta_g \tau^{(1)}}\right).
\end{align}
When $k_{cg} = 1$ and $\beta_g = \gamma_g$, then $u\left(\tau^{(1)}\right)$ 
has the same solution, and $s\left(\tau^{(1)}\right)$ becomes
\begin{align}
s\left(\tau^{(1)}\right)=s_0^{(1)}{ }_g e^{-\gamma_g \tau^{(1)}}
  +\beta_g u_0^{(1)}{ }_g \tau^{(1)} e^{-\beta_g \tau^{(1)}}.
\end{align}

>>>>

We assume the dynamical gene expression is determined by the RNA splicing
process, and infer the unspliced and spliced gene expression level from the
ordinary differential equation (ODEs) proposed in velocyto [@La_Manno2018-lj] and scVelo
[@Bergen2020-pj]
\begin{equation}
  \label{eq-rna}
\begin{align}
\frac{du}{dt} &= \alpha(t) - \beta u, \quad u(0) = u_0 \\
   \frac{ds}{dt} &= \beta u - \gamma s, \quad s(0) = s_0,
\end{align}
\end{equation}
where $u(t), s(t)$ are the unspliced and spliced expression levels of a gene at time $t$ under a transcription rate $\alpha(t)$ with possible temporal dependence, splicing rate $\beta$, and degradation rate $\gamma$. We specify this model to a setting that depends on cell $c$ and gene $g$ as follows:
\begin{equation}
    \label{eq-rna-cg}
\begin{align}
\frac{du_{cg}}{dt} &= \alpha_{cg}(t) - \beta_{g} u_{cg}, \quad u_{cg}(0) = u_{cg}^{(0)} \\
   \frac{ds_{cg}}{dt} &= \beta_{g} u_{cg} - \gamma_{g} s_{cg}, \quad s_{cg}(0) = s_{cg}^{(0)}.
\end{align}
\end{equation}
In the equation, the subscript $c$ is the cell dimension, $g$ is the gene dimension,
$\left( u_{cg}(t), s_{cg}(t) \right)$ 
are the unspliced and spliced expression functions given the
change of time per cell and gene.
We restrict attention to piecewise-constant $\alpha_{cg}(t)$ to capture gene-specific activation and repression. We take special care to model a gene- and cell-specific switching time that marks a single transition from activation to repression by introducing a Bernoulli variable $k_{cg}$ to model unknown transcription state. We assume our cell-by-gene data-matrix arrive as observations of Poisson-counts related to the solution of the above ODEs at unknown times, which is modeled as a relationship between an unknown latent time shared across each cell, $t_c$, and unknown gene-specific time-offsets $t_{0,g}$ where all read counts for a single cell occurred at an unknown, but shared latent time $t_c$. These relative times are also used to parametrize the Bernoulli process for $k_{cg}$. 
Importantly, we recognize that the initial conditions are, in fact, unknown.

@@

Next we introduce the priors for the splicing parameters (where the activation rate $\alpha$ depends on the activation state $k_{cg}$ from above):
\begin{align}
  \alpha^{(0)}_g &\sim \text{LogNormal}(0, 1) \\
  \beta_g &\sim \text{LogNormal}(0, 1) \\
  \gamma_g &\sim \text{LogNormal}(0, 1) \\
  \alpha_{cg} &= \begin{cases}
    \alpha^{(0)}_g & \text{if } k_{cg} = 0 \\
    0 & \text{if } k_{cg} = 1.
  \end{cases}
\end{align}

@@

Next, we define the observation model that gives rise to the observed counts as:
\begin{align}
  \mu^{(u)}_c &= \sum_{g=1}^G {u}^{\text{(obs)}}_{cg}, \quad \mu^{(s)}_c = \sum_{g=1}^G {s}^{\text{(obs)}}_{cg} \\
  \sigma^{(u)}_c &= \sqrt{\frac{1}{G} \sum_{g=1}^G \left( u_{cg}^{\text{(obs)}} - \mu^{(u)}_c \right)^2} \\
  \sigma^{(s)}_c &= \sqrt{\frac{1}{G} \sum_{g=1}^G \left( s_{cg}^{\text{(obs)}} - \mu^{(s)}_c \right)^2} \\
  \eta^{(u)}_c &\sim \text{Normal}\Big(\mu^{(u)}_c, \ \sigma^{(u)}_c\Big) \\
  \eta^{(s)}_c &\sim \text{Normal}\Big(\mu^{(s)}_c, \ \sigma^{(s)}_c\Big) \\
  \hat{\mu}^{(u)}_c &= \sum_{g=1}^G \hat{u}_{cg}, \quad \hat{\mu}^{(s)}_c = \sum_{g=1}^G \hat{s}_{cg} \\
  \lambda^{(u)}_{cg} &= \log(\hat{u}_{cg}) + \log(\eta^{(u)}_{c}) - \log(\hat{\mu}^{(u)}_c) \\
  \lambda^{(s)}_{cg} &= \log(\hat{s}_{cg}) + \log(\eta^{(s)}_{c}) - \log(\hat{\mu}^{(s)}_c) \\
  \hat{u}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(u)}_{cg})\Big) \\
  \hat{s}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(s)}_{cg})\Big)
\end{align}
Here, we use ${u}^{\text{(obs)}}_{cg}, {s}^{\text{(obs)}}_{cg}$ to denote the observed unspliced and spliced counts
for cell $c$ and gene $g$. We use $\hat{u}^{\text{(obs)}}_{cg}, \hat{s}^{\text{(obs)}}_{cg}$ to
denote our generative model's prediction of these unspliced and spliced expression levels.
The generative process for modeling these observed read counts given denoised gene transcript expression level $\hat{u}_{cg}, \hat{s}_{cg}$ considers the expected number of observed reads for a given gene in a given cell as the number of transcripts times the ratio of the cell's total reads to total transcripts.

The linear ODE in Eq. \ref{eq-rna} has an analytic solution under the assumption that the transcription rate $\alpha(t)$ is piece-wise constant, and is derived by the authors of scVelo and a theoretical RNA velocity study
[@Bergen2020-pj;@Li2021-qa].
For brevity, we present the solution under constant transcription rate $\alpha(t) = \alpha_0$, and note that this can be combined to assemble the solution in the piece-wise constant case.
First, let $C = u_0 - \frac{\alpha}{\beta}$. Then, the solution to the ODE in Eq. \ref{eq-rna} is:
\begin{equation}
\label{eq-rna-solution}
\begin{align}
  u(t) &= \frac{\alpha}{\beta} + C e^{-\beta t} \\
  s(t) &= \begin{cases}
    \frac{\alpha}{\gamma} + \frac{\beta}{\gamma - \beta} C e^{-\beta t} + \Big(s_0 - \frac{\alpha}{\gamma} - \frac{\beta}{\gamma - \beta} C\Big) e^{-\gamma t} & \text{if } \beta \neq \gamma \\
    \frac{\alpha}{\beta} + \Big(s_0 - \frac{\alpha}{\beta} + \beta C t\Big) e^{-\beta t} & \text{if } \beta = \gamma.
  \end{cases}
\end{align}
\end{equation}
These solutions are the essential building block used to compute all cell-gene-specific ODE solutions for the unspliced and spliced expression levels in the generative model.
